---
title: "Data Science Project - Loan Approval"
author: "Alex Biuckians, Armin Soroosh, Dania Salman, Kaipu Liu"
output:
  html_document:
    code_folding: hide
    number_sections: false
    theme: journal
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r init, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
library(ezids)
knitr::opts_chunk$set(warning = F, results = "markup", message = F)
# knitr::opts_chunk$set(warning = F, results="hide", message = F)
#knitr::opts_chunk$set(warning = F, message = F)
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```

## Loading the Dataset
```{r read_dataset, echo=TRUE}
# reading dataset
loan_df <- data.frame(read.csv("loan_approval_dataset.csv"))
```

```{r explore_dataset, echo=TRUE}
# exploring dataset
str(loan_df)
```
The dataset has **4269 observations** with **13 variables**

## Data Cleaning & Summary

### subsetting data and converting categorical variables to factor
Removing loan_id column since it only serves as a unique identifier and does not contribute to the analysis. Furthermore, converting several numeric variables into factors, as they represent categorical information rather than continuous numerical values.
```{r data_cleaning, echo=TRUE}
loan_df <- subset(loan_df, select = -c(loan_id))

# converting the numeric variables to factor variables
loan_df$no_of_dependents = as.factor(loan_df$no_of_dependents)
loan_df$education = as.factor(loan_df$education)
loan_df$self_employed = as.factor(loan_df$self_employed)
loan_df$loan_status = as.factor(loan_df$loan_status)
str(loan_df)
```
After conversion, the dataset contains four categorical variables:

- **no_of_dependents:** Number of dependents the applicant has (`6` categories: `0`, `1`, `2`, `3`, `4`, `5`)

- **education:** Applicant’s level of education (`2` categories: `Graduate`, `Not Graduate`)

- **self_employed:** Whether the applicant is self-employed (`2` categories: `Yes`, `No`)

- **loan_status:** Loan approval status — this is the target variable (`2` categories: `Approved`, `Rejected`)

### Finding NA values
```{r, echo=TRUE}
sum(is.na(loan_df))
```
There are no NA values in the data set

### Summary of the dataset
```{r echo=TRUE}
summary(loan_df)
```
Overall, the dataset appears well-structured and balanced. **Categorical variables** such as `education`, `self_employment`, and `loan_status` show nearly **even distributions** across their categories. Most **numerical variables** fall within reasonable ranges, with average loan terms around `11 years` and CIBIL scores centered near `600`, both indicating realistic applicant profiles. However, some financial variables like income, loan amount, and asset values show wide variation and possible outliers. In particular, the presence of negative values in `residential_assets_value` points to potential data quality issues that will need correction before exploration or testing.

### Cleaning up negative values
Checking the number of rows that have negative values for `residential_assets_value` column
```{r echo=TRUE}
sum(loan_df$residential_assets_value < 0)
```
There are `28` entries with negative values.

### Removing entries with negative values
```{r echo=TRUE}
loan_df <- loan_df[loan_df$residential_assets_value >= 0, ]
str(loan_df)
summary(loan_df)
```
After filtering out the rows with negative values, the dataset now has `4,241 observations`. The distributions across all variables look almost identical to before, showing that this cleaning step didn’t affect the data overall.

## Exploratory Data Analysis

### Does education level affect loan approval?
```{r Q1, echo=TRUE}

```

### Do people with higher asset values tend to get approved more often?
```{r Q8, echo=TRUE}

```

### Is there a significant difference in average income between approved and rejected applicants?
```{r Q3, echo=TRUE}

```

### Is there a correlation between applicant's annual income and loan amount requested?
```{r Q4, echo=TRUE}

```

### Is there a significant difference in average CIBIL scores between approved and rejected applicants?
```{r Q5, echo=TRUE}

```

### For rejected loans, is there a significant difference in the mean CIBIL score between applicants with shorter loan term loans (<=10 years) versus longer term loans(>10 years)?
```{r Q6, echo=TRUE}

```

### Is there a correlation between the CIBIL Score and the Loan Term among those whose loans are approved?  
```{r Q7, echo=TRUE}

```

### Does self-employment status affect loan approval?
```{r Q2, echo=TRUE}

```

### Does the number of dependents affect approval of loans?
```{r Q9, echo=TRUE}

```

### Does requested loan term affect approval?
```{r Q10, echo=TRUE}

```




