---
title: "A Data-Driven Study of Loan Approval"
author: "Alex Biuckians, Armin Soroosh, Dania Salman, Kaipu Liu"
bibliography: references.bib
csl: apa.csl
link-citations: true 
output:
  html_document:
    code_folding: hide
    number_sections: false
    theme: journal
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r init, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
library(ezids)
library(dplyr) 
library(tidyr)
knitr::opts_chunk$set(warning = F, results = "markup", message = F)
#knitr::opts_chunk$set(warning = F, results="hide", message = F)
#knitr::opts_chunk$set(warning = F, message = F)
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```
## Introduction
Loan applications continue to rise, creating a growing need for efficient and reliable methods to assess creditworthiness [@Mnkandla2024Loan]. Manual evaluation is slow and inconsistent, so data-driven approaches are essential to support accurate lending decisions [@Anon2025DataDrivenLoanApproval]. To investigate factors influencing loan approval, we use a Kaggle dataset [@sharma2023loan] containing financial and demographic features such as income, employment, asset values, CIBIL score, and loan status. Our exploratory analysis examines how these variables relate to approval outcomes and highlights the most influential features.

Prior research shows that borrowers naturally form clusters based on income, credit history, employment, and demographic attributes (@Wu2018Customer). Motivated by this work, our first modeling question asks whether financial segments exist in this dataset and whether these segments show different loan approval patterns that help explain borrower behaviour. Lenders typically rely on a small set of core indicators such as credit score, income, and loan amount because these variables capture repayment history, repayment capacity, and exposure, and are strong predictors of underwriting decisions and borrower risk (@Bhardwaj2015Credit). Our second modeling question evaluates which combination of these predictors provides the best balance of predictive accuracy and model simplicity.

Together, the exploratory and modeling analyses provide insight into borrower characteristics, identify distinct financial segments, and support more informed, data-driven loan evaluation decisions.

### Loading the Dataset
```{r read_dataset, echo=TRUE}
# reading dataset
loan_df <- data.frame(read.csv("loan_approval_dataset.csv"))
```

```{r explore_dataset, echo=TRUE}
# exploring dataset
str(loan_df)
```
The dataset has **4269 observations** with **13 variables**

## Data Cleaning & Summary

### subsetting data and converting categorical variables to factor
Removing `loan_id` column since it only serves as a unique identifier and does not contribute to the analysis. Furthermore, converting several numeric variables into factors, as they represent categorical information rather than continuous numerical values.
```{r data_cleaning, echo=TRUE}
loan_df <- subset(loan_df, select = -c(loan_id))

# converting the numeric variables to factor variables
loan_df$no_of_dependents = as.factor(loan_df$no_of_dependents)
loan_df$education = as.factor(loan_df$education)
loan_df$self_employed = as.factor(loan_df$self_employed)
loan_df$loan_status = as.factor(loan_df$loan_status)
str(loan_df)
```
After conversion, we have the following variables in the dataset:

#### Variables in dataset
```{r echo=TRUE, results='markup'}
library(knitr)

# Sample data frame
loan_data <- data.frame(
  Variable = c("no_of_dependents", "education", "self_employed", "income_annum", "loan_amount", "loan_term", "cibil_score", "residential_assets_value", "commercial_assets_value", "luxury_assets_value", "bank_asset_value", "loan_status"),
  Description = c("Number of dependents an applicant has ranging from 0 to 5", "Education level (Graduate or Not Graduate)", "Whether the applicant works independently or for an employer", "Annual income of applicant", "Amount of loan requested", "The number of years in which the applicant will repay the loan","Credit score indicating applicants history of repayment", "Value of applicant's residential assets if any", "Value of applicant's commercial assets if any", "Value of applicant's luxury assets if any", "Value of applicant's bank assets if any", "Target variable indicating whether the loan was Approved or Rejected"),
  Type = c("Categorical", "Categorical", "Categorical", "Numeric", "Numeric", "Numeric", "Numeric", "Numeric", "Numeric", "Numeric", "Numeric", "Categorical")
)

# Create table
kable(loan_data)
```

### Finding NA values
```{r, echo=TRUE}
sum(is.na(loan_df))
```
There are no NA values in the data set

### Summary of the dataset
```{r echo=TRUE}
summary(loan_df)
```
Overall, the dataset appears well-structured and balanced. **Categorical variables** such as `education`, `self_employment`, and `loan_status` show nearly **even distributions** across their categories. Most **numerical variables** fall within reasonable ranges, with average loan terms around `11 years` and CIBIL scores centered near `600`, both indicating realistic applicant profiles. However, some financial variables like income, loan amount, and asset values show wide variation and possible outliers. In particular, the presence of negative values in `residential_assets_value` points to potential data quality issues that will need correction before exploration or testing.

### Cleaning up negative values
Checking the number of rows that have negative values for `residential_assets_value` column
```{r echo=TRUE}
sum(loan_df$residential_assets_value < 0)
```
There are `28` entries with negative values.

### Removing entries with negative values
```{r echo=TRUE}
loan_df <- loan_df[loan_df$residential_assets_value >= 0, ]
str(loan_df)
summary(loan_df)
```
After filtering out the rows with negative values, the dataset now has `4,241 observations`. The distributions across all variables look almost identical to before, showing that this cleaning step didn’t affect the data overall.

## Exploratory Data Analysis

### Does education level affect loan approval?
Here, we aim to examine whether the person's **education** (`Graduated` or `Not Graduated`) affects the approval or rejection of a loan. We begin by visualizing the distribution of approved and rejected applications for each education group using a bar chart.
```{r Q1, echo=TRUE}
library(ggplot2)
ggplot(loan_df, aes(x = loan_status, fill = education)) +
  geom_bar(position = "dodge") +
  labs(
    title = "Loan Approval by Education Level",
    x = "Loan Status",
    y = "Number of Applicants"
  ) +
  scale_fill_manual(
    values = c(" Graduate" = "#FF7C61", " Not Graduate" = "#50E5C8")
  ) +
  theme_minimal()
```

Approval rates are nearly **identical** for Graduated and Not Graduated applicants, suggesting education likely doesn’t affect loan approval. A Chi-square test will confirm this statistically.

#### Hypotheses:
- **Null hypothesis (H₀):** Loan approval is independent of education level (Graduated vs Not Graduated).
- **Alternative hypothesis (H₁):** Loan approval depends on education level; there is an association between education and loan approval.

```{r echo=TRUE}
edu_table <- table(loan_df$education, loan_df$loan_status)
print(edu_table)

edu_chi <- chisq.test(edu_table)
print(edu_chi)
```
#### Conclusion
The Chi-square test yielded a **p-value** of `0.8`. Since this is greater than `0.05`, we fail to reject the null hypothesis, indicating that there is no significant association between education level and loan approval. This confirms that, in this dataset, education does not appear to affect the likelihood of a loan being approved.

### Do people with higher asset values tend to get approved more often?
Here, we aim to examine whether the value of a person’s assets, including residential, commercial, luxury, and bank assets, individually affects the approval or rejection of a loan. We begin by visualizing the frequency distribution of each type of asset value.
```{r echo=TRUE}
# 1. Residential Assets Histogram
hist(loan_df$residential_assets_value,
     main = "Distribution of Residential Assets Value",
     xlab = "Residential Assets Value",
     col = "lightblue",
     border = "black")

# 2. Commercial Assets Histogram
hist(loan_df$commercial_assets_value,
     main = "Distribution of Commercial Assets Value",
     xlab = "Commercial Assets Value",
     col = "lightgreen",
     border = "black")

# 3. Luxury Assets Histogram
hist(loan_df$luxury_assets_value,
     main = "Distribution of Luxury Assets Value",
     xlab = "Luxury Assets Value",
     col = "pink",
     border = "black")

# 4. Bank Assets Histogram
hist(loan_df$bank_asset_value,
     main = "Distribution of Bank Assets Value",
     xlab = "Bank Assets Value",
     col = "salmon",
     border = "black")

# Reset plotting layout to default (1 plot per screen) after generation
par(mfrow = c(1, 1))
```
Each type of asset value shows a right-skewed distribution, meaning most individuals hold lower asset values while a smaller number possess much higher ones, indicating non-normality. To statistically verify this, we can apply the Shapiro–Wilk test, which checks whether the data follow a normal distribution.

#### Hypotheses:
- **Null hypothesis (H₀):** The data are normally distributed.
- **Alternative hypothesis (H₁):** The data are not normally distributed.
We use this test to confirm whether the visual observation of skewness is statistically significant.

```{r echo=TRUE}
shapiro.test(loan_df$residential_assets_value)
shapiro.test(loan_df$commercial_assets_value)
shapiro.test(loan_df$luxury_assets_value)
shapiro.test(loan_df$bank_asset_value)
```
Since all **p-values** are far below `0.05`, we **reject the null hypothesis** for each test. This indicates that none of the asset value distributions are normally distributed, which is consistent with the right-skewed patterns observed in the histograms. 
Let's visualize the impact of asset values on loan status using a **box plot**.

```{r echo=TRUE}
# Convert to long format
loan_long <- loan_df %>%
  pivot_longer(
    cols = c(residential_assets_value, commercial_assets_value, luxury_assets_value, bank_asset_value),
    names_to = "asset_type",
    values_to = "asset_value"
  )

# Boxplots all in one figure
ggplot(loan_long, aes(x = asset_type, y = asset_value, fill = loan_status)) +
  geom_boxplot(position = position_dodge(width = 0.8)) +
  labs(
    title = "Asset Distributions by Loan Status",
    x = "Asset Type",
    y = "Asset Value"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c(" Approved" = "#1b9e77", " Rejected" = "#d95f02"))
```
The **median** and **range** of each asset type are similar between `Approved` and `Rejected` loans, suggesting little apparent effect on loan decisions. Since the Shapiro-Wilk test showed that asset values are not normally distributed, we use the Wilcoxon rank-sum test to formally assess differences.

#### Hypotheses for each asset type:
- **Null hypothesis (H₀):** The distribution of asset values is the same for Approved and Rejected loans.
- **Alternative hypothesis (H₁):** The distribution of asset values differs between Approved and Rejected loans.

```{r echo=TRUE}
wilcox.test(residential_assets_value ~ loan_status, data = loan_df)
wilcox.test(commercial_assets_value ~ loan_status, data = loan_df)
wilcox.test(luxury_assets_value ~ loan_status, data = loan_df)
wilcox.test(bank_asset_value ~ loan_status, data = loan_df)
```

#### Conclusion
All **p-values** are greater than `0.05`, so we **fail to reject the null hypothesis** for each asset type. This indicates that there is no statistically significant difference in asset values between Approved and Rejected loans, supporting the initial observation that **asset values do not appear to affect loan approval decisions**.

### Is there a significant difference in average income between approved and rejected applicants?
Here, we aim to examine whether the person's **annual income** affects the approval or rejection of a loan. We begin by visualizing the frequency distribution of annual income.
```{r echo=TRUE}
hist(loan_df$income_annum,
     main = "Distribution of Annual income",
     xlab = "Annual income",
     col = "salmon",
     border = "black")
```
The frequency distribution shows that annual income is **roughly uniform and symmetric**, with **no noticeable skew or extreme outliers**. We can now examine the distribution of annual income for `Approved` and `Rejected` applicants using a box plot.

```{r echo=TRUE}
ggplot(loan_df, aes(x = loan_status, y = income_annum, fill = loan_status)) +
  geom_boxplot() +
  labs(title = "Annual Distribution by Loan Status", x = "Loan Status", y = "Annual Income") +
  scale_fill_manual(values = c(" Approved" = "#12362A", " Rejected" = "#8DD9C1"))
```
The box plot shows that the income distributions for both `Approved` and `Rejected` applicants are quite similar. The **median** income for both groups is approximately `5 million`, and the small difference in the lower quartiles appears negligible. To statistically verify this observation, we will perform a two-sample t-test to compare the means of the two groups.

- **Null Hypothesis (H₀):** There is no significant difference in the average annual income between approved and rejected applicants.

- **Alternative Hypothesis (H₁):** There is a significant difference in the average annual income between approved and rejected applicants.

```{r echo=TRUE}
approved_income <- loan_df$income_annum[loan_df$loan_status == " Approved"]
rejected_income <- loan_df$income_annum[loan_df$loan_status == " Rejected"]
t.test(approved_income, rejected_income)
```

#### Conclusion
Since the `p-value(0.2)> 0.05`, we **fail to reject the null hypothesis**, indicating that there is no statistically significant difference in mean income between Approved and Rejected applicants.So the data supports the earlier observation from the box plot i.e. annual income does not appear to affect loan approval in this dataset.



### Is there a correlation between applicant's annual income and loan amount requested?
Here, we aim to examine whether the **loan amount requested** is influenced by an applicant’s **annual income**. Understanding this relationship helps identify potential **multicollinearity**. Moreover, We already know that annual income is not skewed and roughly uniform and symmetric, we can now visualize the distribution for loan amount requested.
```{r, echo=TRUE}
hist(loan_df$loan_amount,
     main = "Distribution of Loan amount requested",
     xlab = "Loan amount requested",
     col = "yellow",
     border = "black")
```
The distribution of loan amount is right-skewed and deviates from normality. Therefore, we will first visualize the relationship between annual income and the requested loan amount, and then use Spearman’s correlation test, as it does not assume normality.

```{r Q4, echo=TRUE}
annual_income <- loan_df$income_annum
loan_amount <- loan_df$loan_amount

ggplot(data = loan_df, aes(x = loan_amount, y = income_annum)) +
  geom_point(color = "#F7AC19") +
  labs(
    title = "Scatter plot of Annual Income vs Loan Amount",
    x = "Loan Amount",
    y = "Annual Income"
  ) +
  theme_minimal()
```
A positive correlation is evident, indicating that applicants with higher annual incomes tend to apply for and receive larger loan amounts. We can use a spearman correlation test to verify this. 

- **Null Hypothesis (H₀):** There is no relationship between annual income and loan amount among applicants.

- **Alternative Hypothesis (H₁):** Applicants with higher annual incomes tend to apply for and receive larger loan amounts.

```{r echo=TRUE}
cor.test(annual_income, loan_amount, method = "spearman")
```
#### Conclusion:
The Spearman’s rank correlation test shows a very strong positive correlation between annual income and loan amount `(ρ = 0.941, p < 0.001)`. This indicates that as applicants’ annual income increases, the loan amount they apply for and receive also tends to increase. Since the p-value is far below `0.05`, we reject the null hypothesis and conclude that there is a significant positive relationship between annual income and loan amount. This strong association suggests that the two variables convey similar information.

### Is there a significant difference in average CIBIL scores between approved and rejected applicants?
Here, we are investigating to see if there is a significant difference in CIBIL Scores of Approved and Rejected candidates. First we plot the entire distribution to identify outliers then we plot a chart showing the Approved and Rejected populations.
```{r Q5, echo=TRUE}
library(ggplot2)
#CIBIL score frequency distribution
hist(loan_df$cibil_score, main = "Frequency Distribution of CIBIL Scores", xlab = "CIBIL Score", ylab = "Frequency", breaks = 15, col = "#197571")

#CIBIL score frequency box chart
ggplot(loan_df, aes(x = loan_status, y = cibil_score, fill = loan_status)) +
  geom_boxplot() +
  labs(title = "CIBIL Score Distribution by Loan Status", x = "Loan Status", y = "CIBIL Score") +
  scale_fill_manual(values = c(" Approved" = "#197571", " Rejected" = "#FFCF85"))

#t-test
t.test(cibil_score ~ loan_status, data = loan_df)

```

#### Hypotheses:
- **Null hypothesis (H₀):** The average CIBIL Scores for Approved and Rejected applicants are equal.
- **Alternative hypothesis (H₁):** The average CIBIL Scores for Approved and Rejected applicants are not equal.

We use this test to confirm whether the CIBIL Scores for the Approved and Rejected applicants are statistically significantly different or the same.

#### Conclusion
The overall CIBIL score frequency distribution shows no apparent outliers. The individual distributions for approved and rejected loans show a few outliers, but it is not necessary to remove them. Even though the outliers are dragging the two means closer to each other, the means are still significantly different even when including them in the t-test.

There is a significant difference in CIBIL scores between approved and rejected applicants because the p-value, `2e-16`, is much lower than the standard alpha threshold of `0.05`. This allows us to reject the null hypothesis that the two means of approved and rejected applicants are **equal**.

### For rejected loans, is there a significant difference in the mean CIBIL score between applicants with shorter loan term loans (<=10 years) versus longer term loans(>10 years)?
Here, we are investigating to see if there is a significant difference in CIBIL Scores in the Rejected applicant population depending if they want a short (less than or equal to 10 years) vs longer term loan (greater than 10 years). First we plot the entire distribution to identify outliers then we plot a chart showing the Rejected population with shorter and longer term loans.
```{r Q6, echo=TRUE}

#created subsets for shorter and longer loans then ran t-test
rejected_shorter <- subset(loan_df, trimws(loan_status) == "Rejected" & loan_term <= 10)
rejected_longer <- subset(loan_df, trimws(loan_status) == "Rejected" & loan_term > 10)

#Overall loan term frequency distribution
ggplot(loan_df, aes(x = factor(loan_term))) + geom_bar(fill = "#7B3D91") + labs(title = "Frequency Distribution of Loan Terms", x = "Loan Term (Years)", y = "Frequency")

#boxplot
boxplot(
  rejected_shorter$cibil_score,
  rejected_longer$cibil_score,
  names = c("Shorter (<= 10 yrs)", "Longer (> 10 yrs)"),
  main = "CIBIL Score Distribution for Rejected Loans",
  ylab = "CIBIL Score",
  xlab = "Loan Term Group",
  col = c("#A769C2", "#87BDCC")  # add your custom colors here
)
#removing outliers
removeOutliers <- function(x) {Q1 <- quantile(x, 0.25)
Q3 <- quantile(x, 0.75);
IQR_val <- Q3 - Q1;
lower_bound <- Q1 - (1.5 * IQR_val);
upper_bound <- Q3 + (1.5 * IQR_val);
  
return(x[x >= lower_bound & x <= upper_bound]);}

cleaned_shorter_cibil <- removeOutliers(rejected_shorter$cibil_score)
cleaned_longer_cibil <- removeOutliers(rejected_longer$cibil_score)

t.test(rejected_shorter$cibil_score, rejected_longer$cibil_score)
t.test(cleaned_shorter_cibil, cleaned_longer_cibil)

```
#### Hypothesis:
- **Null hypothesis (H₀):** The average CIBIL Scores of Rejected applicants with shorter loan terms (less than or equal to 10 years) is equal to that of Rejected applicants with longer loan term (greater than 10 years).
- **Alternative hypothesis (H₁):** The average CIBIL Scores of Rejected applicants with shorter loan terms (less than or equal to 10 years) is not equal to that of Rejected applicants with longer loan term (greater than 10 years).

We use this test to confirm whether the CIBIL Scores for the Rejected applicants with shorter loan terms and longer loan terms are statistically significantly different or the same.

#### Conclusion
The overall loan terms frequency distribution shows no apparent outliers however there are a few in the shorter and longer loan term groups. Having this said, even when removing these outliers there no significant differences found between the two groups and their respective CIBIL scores.

There is not a significant difference in CIBIL scores between shorter `(<= 10 years)` and longer `(>10 years)` term loans within the rejected group of applicants because the **p-value**, `1`, is much higher than the standard alpha threshold of `0.05`. This allows us to accept the null hypothesis that the two means of shorter and longer term loans of rejected applicants are equal.

### Is there a correlation between the CIBIL Score and the Loan Term among those whose loans are approved?
Here, we are investigating whether or not there is a correlation among Approved applicants between the two variables, CIBIL Score and Loan Term (Years). A correlation test is conducted to determine the correlation coefficient and p-value. A graph is also plotted showing the relationship between CIBIL Score and Loan Term.
```{r Q7, echo=TRUE}

#correlation coefficient
corr_r <- cor(loan_df[trimws(loan_df$loan_status) == "Approved", "cibil_score"], loan_df[trimws(loan_df$loan_status) == "Approved", "loan_term"], method = "pearson")

#correlation test
cor.test(loan_df[trimws(loan_df$loan_status) == "Approved", "cibil_score"], loan_df[trimws(loan_df$loan_status) == "Approved", "loan_term"], method = "pearson")

#scatter plot
ggplot(subset(loan_df, trimws(loan_status) == "Approved"), aes(x = loan_term, y = cibil_score)) + geom_point(alpha = 0.4, color = "#AAE5D8") + geom_smooth(method = "lm", se = FALSE, color = "#061411") + labs(title = "CIBIL Score vs. Loan Term for Approved Loans", x = "Loan Term (Years)", y = "CIBIL Score")

```

#### Hypothesis:

- **Null hypothesis (H₀):** There is no correlation between CIBIL Score and Loan Term (Years) of Approved applicants.
- **Alternative hypothesis (H₁):** There is a correlation between CIBIL Score and Loan Term (Years) of Approved applicants.

We use this test to confirm whether there is a correlation between CIBIL Scores and loan terms for the applicants who were Approved for their loan.

#### Conclusion
The correlation coefficient of `0.21` shows a weak but positive relationship where approved candidates with higher CIBIL scores have a slightly higher tendency to approved for longer loan terms. And the p-value is very small, `2e-16`, meaning that the correlation coefficient is unlikely due to chance.

### Does self-employment status affect loan approval?
Here, we aim to examine whether the person's employment status affects the approval or rejection of a loan. We begin by visualizing the distribution of approved and rejected applications for each employment group using a bar chart.
```{r Q8, echo=TRUE}
ggplot(loan_df, aes(x = self_employed, fill = loan_status)) +
  geom_bar(position = "dodge") +
  labs(
    title = "Loan Approval by self employment status",
    x = "self employment status",
    y = "Number of Applicants"
  ) +
  scale_fill_manual(values = c(" Approved" = "#F2764E", " Rejected" = "#FAF487")) +
  theme_minimal()
```
Approval rates are nearly **identical** for self-employed and non self-employed applicants, suggesting self-employment likely doesn’t affect loan approval. A Chi-square test will confirm this statistically.

#### Hypotheses:
- **Null hypothesis (H₀):** Loan approval is independent of employment status
- **Alternative hypothesis (H₁):** Loan approval depends on employment status; there is an association between employment status and loan approval.

```{r, echo=TRUE}
contable <- table(loan_df$self_employed, loan_df$loan_status)
chisq.test(contable)
```
#### Conclusion
The Chi-square test (`p = 1`) also shows no significant association between self-employment status and loan approval. This confirms that self-employment does not affect the likelihood of loan approval in this dataset.

### Does the number of dependents affect approval of loans?
Here, we will see whether the person's dependents affects the approval or rejection of a loan. We begin by visualizing the distribution of approved and rejected applications for each dependents' group using a bar chart.
```{r Q9, echo=TRUE}
ggplot(loan_df, aes(x = no_of_dependents, fill = loan_status)) +
  geom_bar(position = "dodge") +
  labs(
    title = "Loan Approval by Number of Dependents",
    x = "Number of Dependents",
    y = "Number of Applicants"
  ) +
  scale_fill_manual(values = c(" Approved" = "#0D0942", " Rejected" = "#C4C0F6")) +
  theme_minimal()
```
The bar chart shows that approval and rejection counts are fairly similar across all dependent categories, with no noticeable trend suggesting that the number of dependents strongly influences loan approval outcomes. While minor variations exist, the approval rate appears consistent across groups, indicating that loan approval is likely independent of the number of dependents. A Chi-square test will confirm this statistically.

#### Hypotheses:
- **Null hypothesis (H₀):** Number of dependents is independent of employment status
- **Alternative hypothesis (H₁):** Number of dependents depend on employment status; there is an association between Number of dependents and loan approval.

```{r, echo=TRUE}
contable <- table(loan_df$no_of_dependents, loan_df$loan_status)
chisq.test(contable)
```

#### Conclusion
The Chi-square test (`p = 0.8`) also shows no significant association between `no_of_dependents` and loan approval. This confirms that `no_of_dependents` does not affect the likelihood of loan approval in this dataset.

### Does requested loan term affect approval?
The goal is to see whether loan term will affect the approval of loan, we have already seen the distribution of loan terms so now we will use a box plot to visualize whether certain loan terms have more chance of approval than others.
```{r Q10, echo=TRUE}
ggplot(loan_df, aes(x = loan_status, y = loan_term, fill = loan_status)) +
  geom_boxplot(alpha = 0.6) +
  labs(
    title = "Loan Term by Loan Approval Status",
    x = "Loan Status",
    y = "Loan Term (Years)"
  ) +
  scale_fill_manual(values = c(" Approved" = "#98C25F", " Rejected" = "#EB3BA7")) +
  theme_minimal()
```
The boxplot shows that approved loans have a **wider range** of terms, including very short loans, while rejected loans generally have longer terms. The median loan term for approved loans is `10 years`, compared to `12 years` for rejected loans, indicating that rejected loans tend to have slightly longer terms. To statistically verify this we will perform a t-test with following null and alternate hypothesis:

- **Null Hypothesis (H₀):** There is no significant difference in the mean loan term between approved and rejected loans.
- **Alternate Hypothesis (H₁):** The mean loan term for rejected loans is greater than that for approved loans.


```{r, echo=TRUE}
t.test(loan_term ~ loan_status, data = loan_df, tail="greater")
```
#### Conclusion
The two-sample t-test indicates a significant difference in means (`p < 0.05`) so we can reject the null hypothesis, with approved loans having a shorter average term (`10.4 years`) than rejected loans (`11.7 years`). The `95%` confidence interval for the mean difference is `-1.67 to -0.988` (the interval doesn't have 0), confirming that approved loans tend to have shorter terms.

## Data Modeling

### Q: Do different financial segments show different loan approval patterns, and can these segments help explain which types of borrowers are more likely to be approved or rejected?

#### Filtering numeric columns
```{r echo=TRUE}
data_kmeans <- loan_df[, c("income_annum", "loan_amount", "cibil_score", "residential_assets_value", "commercial_assets_value", "luxury_assets_value", "bank_asset_value")]
cor(data_kmeans)
```

The correlation matrix shows that all asset variables are **positively and strongly related**, so I combined them into a single total asset value to avoid redundancy.

```{r echo=TRUE}
str(loan_df)
loan_df$combined_asset_value <- rowSums(
  loan_df[, c("residential_assets_value",
              "commercial_assets_value",
              "luxury_assets_value",
              "bank_asset_value")],
  na.rm = TRUE
)

data_kmeans <- loan_df[, c("loan_amount", "combined_asset_value", "cibil_score", "income_annum", "loan_status")]
str(data_kmeans)
```

#### Scaling data
```{r echo=TRUE}
data_kmeans_scaled <- scale(
  data_kmeans[, colnames(data_kmeans) != "loan_status"]
)

# this version returns a matrix, so convert to data frame if needed
data_kmeans_scaled <- as.data.frame(data_kmeans_scaled)
```

#### Deciding on optimal number of clusters using elbow plot
```{r echo=TRUE}
wss <- sapply(1:10, function(k){
  kmeans(data_kmeans_scaled, k, nstart = 20)$tot.withinss
})

plot(1:10, wss, type = "b",
     xlab = "Number of clusters (k)",
     ylab = "Total within-cluster sum of squares")
```

The elbow plot indicates a distinct inflection point at `k = 2`, which supports selecting two clusters for the analysis.

#### Run k-means with k=2

```{r echo=TRUE}
set.seed(123) 
k <- 2

km <- kmeans(data_kmeans_scaled, centers = k, nstart = 25)
```

#### Creating cluster profiles

Attaching **cluster labels** to both the scaled and original datasets, then calculating the average unscaled feature values for each cluster to create interpretable cluster profiles.

```{r echo=TRUE}
# df_scaled already contains cluster labels
df_scaled <- as.data.frame(data_kmeans_scaled)
df_scaled$cluster <- factor(km$cluster)

# -----------------------
# 1. Add cluster labels to ORIGINAL data
# -----------------------
df_unscaled <- data_kmeans      # original (unscaled) data
df_unscaled$cluster <- factor(km$cluster)

# -----------------------
# 2. Compute cluster profiles using unscaled values
# -----------------------
cluster_profiles_unscaled <- df_unscaled %>%
  group_by(cluster) %>%
  summarise(
    loan_amount = mean(loan_amount),
    cibil_score = mean(cibil_score),
    income_annum = mean(income_annum),
    combined_asset_value = mean(combined_asset_value)
  )

print(cluster_profiles_unscaled)
```

#### Cluster interpretation
- **Cluster 1: High-value borrowers:** This cluster represents financially strong borrowers who request very large loans and possess significant asset holdings and high annual income. Despite having slightly lower CIBIL scores on average, their high net worth and strong cash flow make them lower risk from a collateral perspective.

- **Cluster 2: Moderate-value borrowers:** This cluster consists of moderate-income, moderate-asset borrowers who apply for smaller loans and have slightly better credit scores. Their financial profile is less substantial than Cluster 1, but their stronger CIBIL scores might indicate better repayment discipline.

#### Calinski-Harabasz Index(Cluster internal assessment)
```{r}
library(fpc)
# calculate CH index
ch <- calinhara(data_kmeans_scaled, km$cluster, cn = 2)
cat("Calinski-Harabasz Index:", ch, "\n")

# compare CH index under different k values
ch_values <- sapply(1:10, function(k) {
  km_temp <- kmeans(data_kmeans_scaled, k, nstart = 20)
  ch <- calinhara(data_kmeans_scaled, km$cluster, cn = k)
})
plot(1:10, ch_values, type = "b", xlab = "k", ylab = "CH Index", main = "CH Index under different K values")
```
We calculated the CH index for the k-means model, which measures the ratio of **variance between clusters to variance within clusters**. Higher values indicate greater separation between clusters and more compact clustering within each group. The CH index for our model was `4531`, a relatively **large value**, suggesting that the clustering result is robust. We then evaluated the CH index across different values of `k`. As shown in the graph, the index reaches its maximum at `k=2`, consistent with the elbow method, confirming that two clusters is the most appropriate choice for this dataset.

### Bootstrap sampling verification(Cluster stability assessment)
```{r}
library(fpc)
set.seed(123)
stab <- clusterboot(data_kmeans_scaled, 
                    clustermethod = kmeansCBI, 
                    k = 2, 
                    runs = 100, 
                    seed = 123)
print(stab$bootmean)
print(stab$bootbrd)  
```
In the bootstrap sampling verification, this model produced very strong results, with bootmean values of `0.994` and `0.995` and bootbrd values of `0` and `0`. These results show that the k-means solution is extremely stable. A bootmean value that is almost equal to `1` indicates that the two clusters appear consistently across more than `99 percent` of the bootstrap samples, which means the model is not creating artificial clusters due to random variation in the data. The bootbrd value measures how uncertain the cluster boundaries are. **A value of zero reflects the highest possible level of boundary clarity**. This confirms that the separation between the two clusters is almost perfectly stable and that the clusters represent **genuine patterns in the data rather than arbitrary divisions created by the algorithm**.

### Calculating Approval percentages for each cluster

```{r echo=TRUE}
approval_percentages <- df_unscaled %>%
  group_by(cluster, loan_status) %>%
  summarise(count = n()) %>%
  mutate(percent = round(count / sum(count) * 100, 2)) %>%
  arrange(cluster)

approval_percentages
```

The clustering analysis reveals two financially distinct borrower segments, but their loan approval patterns differ only slightly. Although the financial profiles of the two groups are clearly separated, their **approval rates remain similar**. The first cluster has an approval rate of `61.4 percent` while the second cluster has an approval rate of `63.1 percent`.

These results show that the **financial segmentation alone does not strongly separate borrowers by approval outcomes**. The clusters highlight meaningful differences in borrowing capacity and asset strength, but these differences do not translate into large differences in approval rates. This suggests that **lenders do not rely primarily on the combined financial structure** captured in the clusters when making decisions. The findings indicate that segmentation is useful for understanding borrower types, but it only partially explains approval behaviour. In later analysis, **CIBIL score emerges as a much stronger driver of approval**, which clarifies why the approval percentages across clusters remain close despite their financial differences.

### Q:  Which set of predictors—1, 2, or 3—provides the optimal balance of predictive power and model complexity for determining loan approval?

The three variables we decided to choose are the `CIBIL score`, `loan amount` and `loan term`. These three would be important because they are vital for the credit risk assessment based on our EDA

- **Null hypothesis (H₀):** All three models are equally effective indicators, or the simplest model is the best since it factors in adding more variables `(AIC1≈AIC2≈AIC3)`  
- **Alternate Hypothesis (H₁):** the model with the minimum AIC is the best indicator, demonstrating a statistically superior balance of fit and model simplicity compared to the other two.

First let us run **Chi-squared test of independence** between predicting and target variables.

```{r}
chisq.test(loan_df$loan_status, loan_df$cibil_score)
chisq.test(loan_df$loan_status, loan_df$loan_amount)
chisq.test(loan_df$loan_status, loan_df$loan_term)
```

These Chi-squared tests on the three variables, `CIBIL score`, `loan amount`, and `loan term`, indicate that the CIBIL score and loan term are the variables with a low p-value (2.2e-16). This means we reject the null hypothesis that that CIBIL score and loan approval are independent and have no relationship. This means that the **CIBIL score is dependent on the loan status** and that the **loan term is significantly associated with the loan status**. We fail to reject the null hypotheses for the loan amount compared to loan status, as the p-values is greater than .05. The loan amount p-value being .8637. This means that the **loan amount and loan approval are independent from each other**.

```{r}
library(regclass)
library(ResourceSelection)

LogitM1 <- glm(loan_status ~ cibil_score, data = loan_df, family = "binomial")
summary(LogitM1)

#evaluation step - confidence interval
confint.default(LogitM1)

#evaluation step - confusion matrix
confusion_matrix(LogitM1)

#evaluation step - Hosmer and Lemeshow
hoslem.test(as.numeric(loan_df$loan_status) - 1, fitted(LogitM1))

#evaluation step - McFadden R^2
null_tLogit <- glm(loan_status ~ 1, data = loan_df, family = "binomial")
mcFadden = 1 - logLik(LogitM1) / logLik(null_tLogit)
cat("McFadden R-squared: ", format(mcFadden, digits=3), "\n")
```

```{r}
LogitM2 <- glm(loan_status ~ cibil_score + loan_term, data = loan_df, family = "binomial")
summary(LogitM2)

#evaluation step - confidence interval
confint.default(LogitM2)

#evaluation step - confusion matrix
confusion_matrix(LogitM2)

#evaluation step - Hosmer and Lemeshow
hoslem.test(as.numeric(loan_df$loan_status) - 1, fitted(LogitM2))

#evaluation step - McFadden R^2
null_tLogit <- glm(loan_status ~ 1, data = loan_df, family = "binomial")
mcFadden = 1 - logLik(LogitM2) / logLik(null_tLogit)
cat("McFadden R-squared: ", format(mcFadden, digits=3), "\n")
```

```{r}
LogitM3 <- glm(loan_status ~ cibil_score + loan_term + loan_amount, data = loan_df, family = "binomial")
summary(LogitM3)

#evaluation step - confidence interval
confint.default(LogitM3)

#evaluation step - confusion matrix
confusion_matrix(LogitM3)

#evaluation step - Hosmer and Lemeshow
hoslem.test(as.numeric(loan_df$loan_status) - 1, fitted(LogitM3))

#evaluation step - McFadden R^2
null_tLogit <- glm(loan_status ~ 1, data = loan_df, family = "binomial")
mcFadden = 1 - logLik(LogitM3) / logLik(null_tLogit)
cat("McFadden R-squared: ", format(mcFadden, digits=3), "\n")
```
  

All three models are highly significant to affect on loan approval, as they all have low p-vaues. `M1`, uses `CIBIL Score` only. `M2` uses `CIBIL score + loan term`. `M3` uses `CIBIL score + loan term + loan_amount`. `M1` has an **R-squared** value of `.623`, an **AIC score** of `2126`,  a **Hosmer and Lemeshow X-squared value** of `623` with a **p-value** of `2e-16`. The **accuracy** on `M1` is `91.77%`. `M2` has an **R-squared** value of `.659`, an **AIC score** of `1925`, and a **Hosmer and Lemeshow X-squared value** of `182` with a **p-value** of `2e-16`. The **accuracy** on `M2` is `91.56%`. `M3` has an **R-squared** value of `.660`, an **AIC score** of `2126`,  and a **Hosmer and Lemeshow X-squared value** of `181` with a **p-value** of `2e-16`. The **accuracy** on `M3` is `91.63%`.

**M1 was the best performing** given the data set. This is due to the fact that it only uses **one predictor**. The simplest model is less prone to overfitting and makes the data easier to interpret.  The accuracy is the highest value as well and the R^2 is marginal. The change of less than 4 percentage points in varaince explained may not justify the added complexity and data requirements. It is not worth adding additional features, we used log reg because the final result (loan approval) is a binary categorical variable, EDA showed cibil score and loan term had strongest relationship with approval so we used those in the log reg models.  To answer the question that is posed, we reject the null hypothesis and state that **Model1 with just the CIBIL score is the best predictor**.

## Overall Findings and Future Work
This study integrated exploratory data analysis and predictive modeling to examine factors influencing loan approval. **The analysis consistently highlights CIBIL score as the primary determinant**, with **loan term also playing a supporting role**. Financial segmentation and other features such as income, assets, and requested loan amount provide additional context but are less critical in driving approval decisions.

The combined insights from clustering and logistic regression confirm that while borrower profiles differ in financial strength, **approval decisions are largely guided by creditworthiness rather than overall asset or income level**. This highlights the value of data-driven approaches in identifying the most meaningful predictors and supporting consistent, objective lending decisions.

Future work can expand on these findings by exploring more sophisticated predictive models, incorporating behavioral or temporal data, and testing whether additional features can improve prediction beyond the dominant effect of CIBIL score.

### References
