---
title: "Data Science Project - Loan Approval"
author: "Alex Biuckians, Armin Soroosh, Dania Salman, Kaipu Liu"
output:
  html_document:
    code_folding: hide
    number_sections: false
    theme: journal
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r init, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
library(ezids)
# knitr::opts_chunk$set(warning = F, results = "markup", message = F)
knitr::opts_chunk$set(warning = F, results="hide", message = F)
#knitr::opts_chunk$set(warning = F, message = F)
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```

## Loading the Dataset
```{r read_dataset, echo=TRUE}
# reading dataset
loan_df <- data.frame(read.csv("loan_approval_dataset.csv"))
```

```{r explore_dataset, echo=TRUE}
# exploring dataset
str(loan_df)
```
The dataset has **4269 observations** with **13 variables**

## Data Cleaning & Summary

### subsetting data and converting categorical variables to factor
Removing loan_id column since it only serves as a unique identifier and does not contribute to the analysis. Furthermore, converting several numeric variables into factors, as they represent categorical information rather than continuous numerical values.
```{r data_cleaning, echo=TRUE}
loan_df <- subset(loan_df, select = -c(loan_id))

# converting the numeric variables to factor variables
loan_df$no_of_dependents = as.factor(loan_df$no_of_dependents)
loan_df$education = as.factor(loan_df$education)
loan_df$self_employed = as.factor(loan_df$self_employed)
loan_df$loan_status = as.factor(loan_df$loan_status)
str(loan_df)
```
After conversion, the dataset contains four categorical variables:

- **no_of_dependents:** Number of dependents the applicant has (`6` categories: `0`, `1`, `2`, `3`, `4`, `5`)

- **education:** Applicant’s level of education (`2` categories: `Graduate`, `Not Graduate`)

- **self_employed:** Whether the applicant is self-employed (`2` categories: `Yes`, `No`)

- **loan_status:** Loan approval status — this is the target variable (`2` categories: `Approved`, `Rejected`)

### Finding NA values
```{r, echo=TRUE}
sum(is.na(loan_df))
```
There are no NA values in the data set

### Summary of the dataset
```{r echo=TRUE}
summary(loan_df)
```
Overall, the dataset appears well-structured and balanced. **Categorical variables** such as `education`, `self_employment`, and `loan_status` show nearly **even distributions** across their categories. Most **numerical variables** fall within reasonable ranges, with average loan terms around `11 years` and CIBIL scores centered near `600`, both indicating realistic applicant profiles. However, some financial variables like income, loan amount, and asset values show wide variation and possible outliers. In particular, the presence of negative values in `residential_assets_value` points to potential data quality issues that will need correction before exploration or testing.

### Cleaning up negative values
Checking the number of rows that have negative values for `residential_assets_value` column
```{r echo=TRUE}
sum(loan_df$residential_assets_value < 0)
```
There are `28` entries with negative values.

### Removing entries with negative values
```{r echo=TRUE}
loan_df <- loan_df[loan_df$residential_assets_value >= 0, ]
str(loan_df)
summary(loan_df)
```
After filtering out the rows with negative values, the dataset now has `4,241 observations`. The distributions across all variables look almost identical to before, showing that this cleaning step didn’t affect the data overall.

## Exploratory Data Analysis

### Does education level affect loan approval?
```{r Q1, echo=TRUE}

```

### Do people with higher asset values tend to get approved more often?
```{r Q2, echo=TRUE}

```

### Is there a significant difference in average income between approved and rejected applicants?
```{r Q3, echo=TRUE}
# Group by loan approval status and extract their annual income data
approved_income <- loan_df$income_annum[loan_df$loan_status == " Approved"]
rejected_income <- loan_df$income_annum[loan_df$loan_status == " Rejected"]
# Calculate each group's mean annual income
mean_approved <- mean(approved_income, na.rm = TRUE)
mean_rejected <- mean(rejected_income, na.rm = TRUE)
cat("Mean annual income of Approved group:", mean_approved, "\n")
cat("Mean annual income of Rejected group:", mean_rejected, "\n")

# Conduct hypothesis testing
t_test_result <- t.test(approved_income, rejected_income)
print("t-test result：")
print(t_test_result)
# Draw the histogram
hist(approved_income, 
     col = "lightgreen",
     main = "The distribution of annual income", 
     xlab = "Annual income", 
     ylab = "Frequency",
     breaks = 30)
hist(rejected_income, 
     col = "red", 
     breaks = 30,
     add = TRUE)
legend("topright", 
       legend = c("Approved", "Rejected"), 
       fill = c("lightgreen", "red")) 

```
The average income of the approved group is 5,025,904, and that of the rejected group is 5,113,825, with a difference of approximately 87,921. Then I did the hypothesis testing to do advanced judgment.

H0: There is no significant difference in the average annual income between two groups.
H1: There is a significant difference in the average annual income between two groups.

According to the result of t-test:
t-value: -1, indicating that the difference between the two group means is not prominent compared to the sample variation.
p-value: 0.3 which is much larger than the significance level of 0.05, indicating that the different is non-significant.
95% confidence interval: [-260,821, 84,978]. The interval contains 0, indicating that the mean difference between the two groups may be 0.
Finally, I drew the histogram chart, the distribution of each group's annual income is uniform. Also enhanced the conclusion.

All above suggests that income alone may not be a strong predictor of loan approval in this dataset.



### Is there a correlation between applicant's annual income and loan amount requested?
```{r Q4, echo=TRUE}
annual_income <- loan_df$income_annum
loan_amount <- loan_df$loan_amount

# Check whether normal distribution
shapiro_income <- shapiro.test(annual_income)
shapiro_loan <- shapiro.test(loan_amount)
cat("p-value of annual income：", shapiro_income$p.value, "\n")
cat("p-value of loan amount：", shapiro_loan$p.value, "\n")
# Choose the correlation coefficient
if (shapiro_income$p.value > 0.05 & shapiro_loan$p.value > 0.05) {
  cor_result <- cor.test(annual_income, loan_amount, method = "pearson")
  cat("Pearson result：\n")
} else {
  cor_result <- cor.test(annual_income, loan_amount, method = "spearman")
  cat("Spearman reslut：\n")
}
print(cor_result)

# Draw the plot
plot(
  x = annual_income,
  y = loan_amount,
  main = "Annual Income vs. Loan Amount Requested",
  xlab = "Annual Income",
  ylab = "Loan Amount Requested",
  pch = 16,
  col = "lightblue",
  cex = 0.8
)
# Add the regression line
abline(lm(loan_amount ~ annual_income), col = "red", lwd = 2)
legend("topleft", 
       legend = "regression line", 
       col = "red", 
       lty = 1, 
       lwd = 2)




```
I checked whether the data of annual income and loan amount is follow the normal distribution and choose the Spearman  correlation coefficient according to the result.
The rho is 0.941 which is close to 1, indicating that there is a very strong positive monotonic correlation between the annual income and loan amount exist and the p-value is far less than 0.05 also showed this correlation is very significant in statistic.
Then according to the plot chart, it's very clear that the loan amount shows an obvious upward trend from the bottom left to the top right as annual income increases, and the red regression line further strengthens the visual evidence of this linear association, indicating a strong positive correlation tendency between the two.
In short, this analysis clearly draws the conclusion that "annual income is a key factor affecting the loan application amount, and there is a strong positive correlation between the two."

### Is there a significant difference in average CIBIL scores between approved and rejected applicants?
```{r Q5, echo=TRUE}

```

### For rejected loans, is there a significant difference in the mean CIBIL score between applicants with shorter loan term loans (<=10 years) versus longer term loans(>10 years)?
```{r Q6, echo=TRUE}

```

### Is there a correlation between the CIBIL Score and the Loan Term among those whose loans are approved?  
```{r Q7, echo=TRUE}

```

### Does self-employment status affect loan approval?
```{r Q8, echo=TRUE}

```

### Does the number of dependents affect approval of loans?
```{r Q9, echo=TRUE}

```

### Does requested loan term affect approval?
```{r Q10, echo=TRUE}

```




